---
title: "Multiple testing with mtp"
author: Gwenaël G.R. Leday
output: rmarkdown::html_vignette
bibliography: refs.bib
biblio-style: "apalike"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Multiple testing with mtp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

## Multiple hypothesis testing

In multiple hypothesis testing, $m > 1$
null hypotheses $H_{1}, \ldots, H_{m}$ with
corresponding p-values $p_{1}, \ldots, p_{m}$
are tested simultaneously.

## Why using multiple testing procedures?

When conducting a single statistical test ($m = 1$),
the probability of committing an error when
rejecting the null hypothesis (a Type I error)
is bounded by the significance level $\alpha$,
which is conventionally set at 0.05.

However, when multiple tests are performed,
the risk of committing such errors increases with
the number of tests, and with a large number of
test these errors become almost inevitable.

Multiple testing procedures address this issue
by providing statistical control over the
occurrence of Type I errors across all tests,
using appropriate error criteria.

They achieve this by adjusting the significance level of
individual tests so that the overall occurrence of
Type I errors remains bounded according to the chosen
error criterion and pre-specified level.

Different criteria offer different forms of control,
and multiple adjustment methods are available for
each criterion.

In essence, multiple testing procedures:

- provide statistical guarantees about the extent of
Type I errors (false positives)

- help strike a balance between limiting Type I errors
and maintaining statistical power (the ability to
correctly reject null hypotheses)

## Error criteria

> **Notation** <br>
$\text{V}$: number of false rejections <br>
$\text{R}$: number of rejected null hypotheses

mtp allows to control Type I errors using:

* the **family-wise error rate** (FWER),
defined as the probability that the number of
false rejections exceeds a predefined
number $k$, with $0 \leq \text{k} < \text{m}$:
$$\text{FWER(k)} = \text{Pr}(\text{V} > \text{k}).$$

* the **false rejection exceedance** (FRX),
defined as the probability that the proportion
of false rejections among all rejected hypotheses
exceeds a predefined proportion $d$, with $0 \leq \text{d} < 1$:
$$\text{FRX(d)} = \text{Pr}(\text{V}/\text{R} > \text{d}).$$

* the **per family error rate** (PFER),
defined as the expected number of false rejections:
$$\text{PFER} = \text{E}(\text{V}).$$

* the **false rejection rate** (FRR),
defined as the expected proportion
of false rejections among all rejected hypotheses:
$$\text{FRR} = \text{E}(\text{V}/\text{R}).$$

> **Note** <br>
> The term *discovery* often adopted in literature
to name error criteria is here replaced by the more
neutral term *rejection*. Indeed, not all rejections
result in a "discovery" (e.g. equivalence tests).


## Adjustment methods

To control Type I errors according to the selected criterion,
the significance level of each test must be adjusted appropriately.
This adjustment involves either increasing the p-value
or, equivalently, decreasing the critical value
with an **adjustment factor**.

Hence, adjustments methods provide either of the following
(from which final rejection decisions can be made):

* the **adjusted p-values**
$\widetilde{p}_{1}, \ldots, \widetilde{p}_{m}$

* the **adjusted critical values**
$\widetilde{\alpha}_{1}, \ldots, \widetilde{\alpha}_{m}$

For a given criterion different adjustment methods are available.
These can be distinguished based on their type:

* **single-step** methods (e.g. Bonferroni's) adjust each test
regardless of the outcome of other tests.

* **step-down** methods (e.g. Holm's) adjust each test
sequentially, starting with the most significant
(with smallest p-value).

* **step-up** methods (e.g. Hochberg's) adjust each test
sequentially, starting with the least significant
(with largest p-value).

> **Note:**
> The code used for the step-down and
step-up procedures in mtp is largely based
on the code of `p.adjust()` from package stat.

## Decision procedure

After adjustment, the **decision procedure** consists in
applying one of the following (equivalent) rules:

* reject null hypotheses with <u>adjusted p-values</u> smaller than or
equal to the <u>critical value</u> $\alpha$, i.e.:
$$
 \displaystyle{
  \text{reject}\ \ H_{j} \qquad
  \text{if} \qquad \widetilde{p}_{j} \leq \alpha,
  \qquad\text{for}\ j=1, \ldots, m.
 }
$$

* reject null hypotheses with <u>observed p-values</u> smaller than or
equal to the <u>adjusted critical values</u>, i.e.:
$$
 \displaystyle{
 \text{reject}\ \ H_{j} \qquad
  \text{if} \qquad p_{j} \leq \widetilde{\alpha}_{j},
  \qquad\text{for}\ j=1, \ldots, m.
 }
$$

# The mtp package

The following table lists the core functions of mtp.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| html-table-processing: none
#| results: "asis"

# function list for FWER
funs_fwer <- data.frame(
  Function = c(
    "`fwer_bon()`",
    "`fwer_holm()`",
    "`fwer_hoch()`",
    "`fwer_bon_a()`",
    "`fwer_bon_w()`"
  ),
  Procedure = c(
    "generalized Bonferroni",
    "generalized Holm",
    "generalized Hochberg",
    "generalized adaptive Bonferroni",
    "generalized weighted Bonferroni"
  ),
  Type = c(
    "single-step",
    "step-down",
    "step-up",
    "single-step",
    "single-step"
  ),
  Error = c(
    rep("FWER(k)", 5)
  )
)

# function list for FRX
funs_frx <- data.frame(
  Function = c(
    "`frx_holm()`",
    "`frx_hoch()`"
  ),
  Procedure = c(
    "Lehmann & Romano's modified Holm",
    "Guo's modified Hochberg"
  ),
  Type = c(
    "step-down",
    "step-up"
  ),
  Error = c(
    rep("FRX(d)", 2)
  )
)

# function list for FRR
funs_frr <- data.frame(
  Function = c(
    "`frr_bh()`",
    "`frr_bh_a()`"
  ),
  Procedure = c(
    "Benjamini-Hochberg",
    "adaptive Benjamini-Hochberg"
  ),
  Type = c(
    "step-up",
    "step-up"
  ),
  Error = c(
    rep("FRR", 2)
  )
)

# combine function lists
funs <- rbind(
  funs_fwer,
  funs_frx,
  funs_frr
)

# make table
knitr::kable(funs) |>
  kableExtra::collapse_rows(
    columns = 4,
    valign = "middle"
  ) |>
  kableExtra::row_spec(
    row = nrow(funs_fwer),
    hline_after = TRUE
  )
```

These functions allow to obtain either of the following:

* the **adjusted p-values**
when the function argument `output = "p_values"`.

* the **adjusted critical values**
when the function argument `output = "c_values"`.

* the **rejection decisions**
when the function argument `output = "decisions"`.

* the **adjustment factors**
when the function argument `output = "factors"`.

Details on how adjusted p-values, critical values and
adjustment factors are computed are provided
in each function's documentation.

# Alternative packages

If mtp does not meet your needs,
perhaps one of the following packages will:

* [cherry](https://cran.r-project.org/package=cherry):
Multiple Testing Methods for Exploratory Research
* [DiscreteFDR](https://cran.r-project.org/package=DiscreteFDR):
FDR Based Multiple Testing Procedures with Adaptation for Discrete Tests
* [DiscreteFWER](https://cran.r-project.org/package=DiscreteFWER):
FWER-Based Multiple Testing Procedures with Adaptation for Discrete Tests
* [elitism](https://cran.r-project.org/package=elitism):
Equipment for Logarithmic and Linear Time Stepwise Multiple Hypothesis Testing
* [FAMT](https://cran.r-project.org/package=FAMT):
Factor Analysis for Multiple Testing
* [fdrci](https://cran.r-project.org/package=fdrci):
Permutation-Based FDR Point and Confidence Interval Estimation]
* [fdrtool](https://cran.r-project.org/package=fdrtool):
Estimation of (Local) False Discovery Rates and Higher Criticism
* [flip](https://cran.r-project.org/package=flip):
Multivariate Permutation Tests
* [FixSeqMTP](https://cran.r-project.org/package=FixSeqMTP):
Fixed Sequence Multiple Testing Procedures
* [pwrFDR](https://cran.r-project.org/package=pwrFDR):
FDR Power
* [FDX](https://cran.r-project.org/package=FDX):
False Discovery Exceedance Controlling Multiple Testing Procedures
* [gMCP](https://cran.r-project.org/package=gMCP):
Graph Based Multiple Comparison Procedures
* [graphicalMCP](https://cran.r-project.org/package=graphicalMCP):
Graphical Multiple Comparison Procedures
* [hommel](https://cran.r-project.org/package=hommel):
Methods for Closed Testing with Simes Inequality, in Particular Hommel's Method
* [MHTdiscrete](https://cran.r-project.org/package=MHTdiscrete):
Multiple Hypotheses Testing for Discrete Data
* [MHTmult](https://cran.r-project.org/package=MHTmult):
Multiple Hypotheses Testing for Multiple Families/Groups Structure
* [multcomp](https://cran.r-project.org/package=multcomp):
Simultaneous Inference in General Parametric Models
* [multtest](https://bioconductor.org/packages/multtest/):
Resampling-based multiple hypothesis testing
* [multxpert](https://cran.r-project.org/package=multxpert):
Common Multiple Testing Procedures and Gatekeeping Procedures
* [mutoss](https://cran.r-project.org/package=mutoss):
Unified Multiple Testing Procedures
* [nparcomp](https://cran.r-project.org/package=nparcomp):
Multiple Comparisons and Simultaneous Confidence Intervals
* [qvalue](https://bioconductor.org/packages/qvalue/):
Q-value estimation for false discovery rate control
* [sgof](https://cran.r-project.org/package=sgof):
Multiple Hypothesis Testing
* [someMTP](https://cran.r-project.org/package=someMTP):
Some Multiple Testing Procedures
* [StepwiseTest](https://cran.r-project.org/package=StepwiseTest):
Multiple Testing Method to Control Generalized Family-Wise Error Rate and False Discovery Proportion


# References

---
nocite: '@*'
---
